The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-13 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{K}(h)}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=\min)$
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition}\ref{apx:kalman}.

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
  \end{figure}
\end{example}

\paragraph{Kryptotypic variation and hybrid incompatibility}
Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.

For instance, diploid system coefficients are derived by taking the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes ($\Sys$ and $\Sys'$) have the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.

If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

\paragraph*{Hybrid phenotypic breakdown between oscillators}
For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid $\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$ fails to oscillate and exhibits qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
    }
  \end{figure}
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

@@

\plr{Concentrations are "logarithmic", you mean that $u_1(t) = \log(\text{concentration of molcule 1 at time t})$.}
Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), 
with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). 
Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). 
The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy. 
\plr{I think you argue below that the optimal response is a complete switch from one enzyme to the other in the presence of the second sugar? Should say so here.}
That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.

Minimally such a system may have the architecture,
\begin{align*}
    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
\end{align*}
$\mathcal{S}_{\min}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.

The impulse response matrix of this system is, 
\begin{align*}
    h(t) = \begin{bmatrix} 1 & - t \\ 0 & 1 \end{bmatrix}  .
\end{align*}

The phenotype is, 
\begin{align*}
  \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} h(s-t) u(s) ds
\end{align*}
($\kappa(0)$ can be set to something like $\left[ -10, -10 \right]$, assuming the transient transcription factor and enzyme concentrations in the organism are typically quite low).

We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, 
as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions 
-- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. 
\plr{Ah -- this is much simpler if we let $B$ and $C$ change as well!}
This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

\plr{where did $\widehat{\mathcal{S}}$ come from? need a lead-in here.}
$\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  
\begin{align*}
    \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
      \widehat{A} = V \begin{bmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{bmatrix} V^{-1}; \quad \widehat{B} = V \begin{bmatrix} X_{3} \\ B_{\min} \end{bmatrix}; \quad \widehat{C} = \begin{bmatrix} 0 & C_{\min} \end{bmatrix} V^{-1} \end{array} \right\}
\end{align*}
where $V$ is any invertible $4 \times 4$ matrix and $X_{j}$ is any $2 \times 2$ matrix.

\plr{the point here is not clear.}
So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
 \begin{align*}
   A' &= \begin{bmatrix}
     1.6923  &  1.5385 &  -2.6154 &  -2 \\
     0.8462  &  0.7692 &  -1.3077 &  -1 \\
     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
     0.4231  &  0.3846 &  -0.6538 &  -0.5
   \end{bmatrix} \\
   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
 \end{align*}

Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
any $n$-dimensional system can be constructed using this method 
-- applying a change of coordinates to the Kalman decomposition 
-- to construct a mechanistically different system with identical phenotypic dynamics. 
Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architectures to be biologically appropriate. 
\plr{what are you thinking of as being a potential problem here? say so explicitly.}

@@

\section{Sexual reproduction in linear systems}\label{apx:reproduction}

  Recombination: gamete element $A''$ is formed by recombining (randomly swapping rows) between two parental, possibly distinct, system elements $A$ and $A'$ as follows,
  \begin{align*}
    A'' = MA + (I-M)A'
  \end{align*}
  where $M$ is a diagonal matrix with only $0$s and $1$s on the diagonal.

>>>>

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{K}}
%\newcommand{\allS}{\mathcal{A}}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VA_0V^{-1}, VB_0, C_0V )$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
  \js{what do you like better for $\allS$? $\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$? Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output. A minimal system's components are both reachable and observable. In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs. Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response. Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components. After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton. 
Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question. 
\js{should the sentence of evolution of nearly minimal be in the paper? Or should we save this for our next paper?}

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    %Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \caption{Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
  \end{figure}
Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
Further suppose that the second sugar is prefered,
%and that the organism has a regulatory network that can deploy situation specific metabolic strategy.
and that
depending on $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 
Furthermore, assume this system contains at least two transcription factors, whose log concentrations are $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.
Minimally such a system may have the architecture,
%\begin{align*}
%    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
%    \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
%\end{align*}
  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right], \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]U^{-1})$. 
%The impulse response of the system is,
%$h(t) = \left[\begin{smallmatrix} 1 & - t \\ 0 & 1 \end{smallmatrix}\right]$.
    We can find alternative equivalent systems by changning coordinates ($U \rightarrow U'$) or, more generally, by applying the Kalman decomposition \ref{apx:kalman}. To illustrate that phenotypic invariance does not require dimensional invariance, we apply the Kalman decomposition to $\Sys_{\min}$ to find 
    %This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion. 
  $\Sys(\small X_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
     V \! \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A \end{smallmatrix}\right] \! V^{-1}, V \! \left[\begin{smallmatrix} X_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \! V^{-1} \end{array} \! \right)
$, both of which are in $\allS$
%  \begin{align*}
%    \Sys(X_{1-3},V) = \left( \begin{array}{ll}
%     V \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{smallmatrix}\right] V^{-1}, V \left[\begin{smallmatrix} X_{3} \\ B_{\min} \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C_{\min} \end{smallmatrix}\right] V^{-1} \end{array} \right)
%    \end{align*}
  ($V$ can be any $4$-dimensional and invertible matrix, $X_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).
%So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
% \begin{align*}
%   A' &= \begin{bmatrix}
%     1.6923  &  1.5385 &  -2.6154 &  -2 \\
%     0.8462  &  0.7692 &  -1.3077 &  -1 \\
%     1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
%     0.4231  &  0.3846 &  -0.6538 &  -0.5
%   \end{bmatrix} \\
%   B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
%     C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
%     -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
% \end{align*}
%Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, 
%any $m$-dimensional system can be constructed using this method 
%-- applying a change of coordinates to the Kalman decomposition 
%-- to construct a mechanistically different system with identical phenotypic dynamics. 
%When applying the Kalman decomposition to real biological networks, one may have to restrict the free parameters, as physiology and biochemistry might not allow all mathematically possible coefficients. We note that as there are so many degrees of freedom, it may be difficult to infer system function solely from structure. 

\paragraph{Sexual recombination and hybrid incompatibility}
Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.

    For instance, the diploid $\{\Sys, \Sys' \}$ has system architure $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}). Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).

Even if two systems yield the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.

If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes (Figure \ref{fig:expF1}).

%\paragraph*{Hybrid phenotypic breakdown between oscillators}
%For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
  \begin{figure}[H] 
    \centering
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross from Example \ref{ex:all_osc}. The $F_1$ hybrid
%$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
gene expression fails to oscillate and instead grows exponentialy,
exhibiting qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
    }
    \label{fig:expF1}
  \end{figure}
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

@@

@@

\section{Meiotic recombination in linear systems}\label{apx:recombination}

  Recombination is performed by taking two analogous system components from $\Sys$ and $\Sys'$ and randomly swapping rows or columns.
  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining (randomly swapping rows or columns) between two, possibly distinct, systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$ such that,
  \begin{align*}
    \Sys'' = \left( \begin{array}{ll}
    A'' &= MA + (I-M)A' , \\
    B'' &= MB + (I-M)B' , \\
    C'' &= CM + C'(I-M)
    \end{array} \right)
  \end{align*}
  where $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$). If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).
